Fluent statistical computing interfaces for biological data analysis

### 2020-07-17 Rmd/00-b-abstract.Rmd
first pass at abstract

Exploratory data analysis  (EDA) is a vital element of the modern 
statistical workflow - it is an analyst’s first pass at 
understanding their data; revealing all it’s messes and (possibly) uncovering hidden insights. It is an iterative process that can 
lead to new hypotheses that can be tested and formalised using 
modelling. However, as data grows in complexity and there is an 
increase in the number of observations, and variables EDA becomes 
increasingly difficult.  This is certainly true in bioinformatic 
data analysis, where there are constraints computation time and 
memory, as well as a lack of guidelines or tooling for performing 
EDA.

This thesis presents a set of tools for performing exploratory 
data analysis in the context of genomics and transcriptomics.  The first tool, “plyranges” provides a domain specific language for 
reasoning about range based genomics data that simplifies common 
operations such as overlaps and aggregations.  Next I extend 
“plyranges”  to develop “superintronic” which presents an 
approach for discovering interesting genomic regions within an 
experimental design. We apply this tool to explore intron signal 
in transcriptomic data. Finally, I turn my attention to focus on 
embedding methods currently used for visualisation of large data 
sets…

>>>>

Exploratory data analysis is vital to modern science workflows; it allows
scientists to grasp problems with their data and generate new hypotheses.
This work explores three facets of exploratory data analysis workflows as 
applied to biological data science: data wrangling, integration and 
visualisation.  It contributes new statistical computing interfaces and 
frameworks with the explicit aim of enabling scientists to understand
their data and models in their biological context. In chapter 
\@ref(ch:plyranges) we show that genomics data can be represented using
tidy data semantics, and consequently the process of wrangling it can
be simplified via our grammar of genomic data transformation. 
The next contribution is exploring the implications
of our grammar on the integration and representation of genomics data.
In chapter \@ref(ch:fluentGenomics), we provide a framework for integrating
genomics data from multiple assays, via combining model estimates over
their genomic regions. Next we extend our grammar to represent single
variable measurements along the genome in multiple ways; 
in chapter \@ref(ch:intron) we present a software tool that allows coverage
scores to be aggregated and visualised over an experimental design and 
genomic features and use this tool to uncover intron signal in RNA-seq data.
Finally, in chapter \@ref(ch:tsne) we contribute a new visualisation 
interface that provides scientists with a toolkit for discovering structure in 
their high dimensional data, and assist them in understanding when non-linear 
dimension reduction has worked appropriately.
